Drug sensitivity prediction with normal inverse Gaussian shrinkage informed by external data

### 2020-03-04 docs/manuscript1.Rnw
commit bfeore push

	\subsection{Bayesian prior model}\label{sec:prior}
	We carry out inference by endowing the parameters with the following priors:
	\begin{subequations}\label{eq:prior}
		\begin{align}
		  \beta_{jd} | \gamma^2_{jd}, \tau_d^2, \sigma^2_d & \sim \mathcal{N}_{p} 
		  (0, \gamma_{jd}^2 \tau_d^2 \sigma_d^2), 
		  \label{eq:betaprior}\\
		  \gamma_{jd}^{2} & \sim \mathcal{IG}(\phi_{jd}, \lambda_{\text{feat}}), \\
		  \tau_d^{2} & \sim \mathcal{IG}(\chi_d, \lambda_{\text{drug}}), 
		  \label{eq:tauprior}\\
		  \sigma_d^{2} & \sim 1/\sigma_d^{3},
		\end{align}
	\end{subequations}
	where $\mathcal{IG}(\phi, \lambda)$ denotes an inverse Gaussian distribution 
	with mean $\phi$ and shape $\lambda$.

	In model (\ref{eq:prior}), $\gamma_{jd}^2$ denotes a local variance 
	component that is supposed to capture local, feature-specific variation in the
	model parameters $\beta_{jd}$, while the global variance component $\tau$ 
	captures the, drug-specific, general trend in $\bm{\beta}_{d}$. Each drug is
	endowed with a random error variance $\sigma_d^2$.

@@

<<dens_kappa, fig.cap="Implied prior densities $\\pi(\\kappa_{jd})$ for the (a) NIG, (b) Student's $t$, and (c) lasso priors. Different line types correspond to different hyperparameter settings. The hyperparameter settings were chosen to show some possible, distinct shapes that each of the priors can take.", out.width="100%", fig.asp=1/3>>=
@

  Figure \ref{fig:dens_kappa} shows that, depending on the choice
  of hyperparameters, the NIG prior can behave similarly to the Student's $t$ 
  prior (decreasing form zero, with substantial mass close to 0), 
  but also rather differently. Our argumentation to model the 
  $\gamma^2_{jd}$ by an inverse Gaussian 
  distribution, as has been suggested in \cite{fabrizi_specification_2016} and 
  \cite{caron_sparse_2008}, is three-fold: (i) the NIG model is more flexible 
  than the lasso prior (as seen from Figure 
  \ref{fig:dens_kappa}), (ii) the NIG prior allows to model
  the means of the $\gamma_{jd}^2$ ($\phi_{jd}$) and $\tau_{jd}^2$ 
  ($\chi_{d}$) as a function of external data 
  more conveniently than the Student's $t$ prior, as explained in Section 
  \ref{sec:empiricalbayes}, and (iii) like the horseshoe, the NIG shrinkage
  weights prior can put mass both near 0 and 1, a desirable property of 
  shrinkage priors.

@@

	A pre-processing step selects 100 features with largest variance, while the 
	168 drug sensitivities for 482 cell lines are 
	simulated from model (\ref{eq:likelihood}) and 
	(\ref{eq:prior}). We draw the error
	variances as $\forall d: \sigma_d^2 \sim \Gamma^{-1}(3,2)$,
	such that the prior $\sigma_d^2$ mean and variance are both one. 
	We consider the following three scenarios for the simulation of the 
	drug and feature variance components:
	\begin{itemize}
	  \item{Scenario 1 fixes $\forall d: \tau_d^2=1$ and draws the 
	    $\gamma_{jd}^2$ according to model (\ref{eq:prior}). We
	    create four external dummy feature covariates that code for four 
	    randomly assigned approximately equally sized groups of features. We set
	    $\bm{\alpha}_{\text{feat}}$ such that the $\gamma_{jd}^2$ of the four
	    groups of features have prior means $\phi_{jd} \in \{ 1, 1/2, 1/4, 1/8 \}$
	    ($\bm{\alpha}_{\text{feat}}=\begin{bmatrix} 1 & 1 & 3 & 7 \end{bmatrix} 
	    \tr$). The prior scale parameter is set to $\lambda_{\text{feat}}=1$}.
	  \item{Scenario 2 fixes $\forall j,d: \gamma_{jd}^2=1$ and draws the 
	    $\tau_d^2$ according to model (\ref{eq:prior}), following a procedure
	    similar to the procedure for the $\gamma_{jd}^2$ in Scenario 1: we 
	    randomly create four groups of drugs with corresponding external drug 
	    dummy variables and set $\bm{\alpha}_{\text{drug}}=\begin{bmatrix} 
	    1 & 1 & 3 & 7 \end{bmatrix} \tr$, such that we have $\chi_{d} \in 
	    \{ 1, 1/2, 1/4, 1/8 \}$. The scale is set to $\lambda_{\text{drug}}=1$.}
	  \item{Scenario 3 combines the procedures from Scenarios 1 and 2 to draw both
	    the $\gamma_{jd}^2$ and $\tau_d^2$ according to (\ref{eq:prior}).}
	\end{itemize}
	We estimate four models: (i) the NIG model that only includes an intercept in
	the external covariates, called NIG$_{\text{f}}^-$, NIG$_{\text{d}}^-$, or
	NIG$_{\text{f}+\text{d}}^-$, depending on which variance components are 
	estimated (feature, drug, or both in Scenarios 1, 2, and 3, respectively),
	(ii) the NIG model estimated as in Section \ref{sec:estimation} that includes
	all external covariates, called NIG$_{\text{f}}$, NIG$_{\text{d}}$, or
	NIG$_{\text{f}+\text{d}}$, again depending on which variance components are 
	estimated, and regular (iii) lasso and (iv) ridge models.
	Exclusion of the external covariates as in the NIG$_{\text{f}}^-$, 
	NIG$_{\text{d}}^-$, and NIG$_{\text{f}+\text{d}}^-$ models
	amounts to direct estimation of common expected 
	prior means $\phi$ and/or $\chi$, instead of regression estimates for the
	$\phi_{jd}$ and/or $\chi_d$ as in the 
	NIG$_{\text{f}}$, NIG$_{\text{d}}$, and NIG$_{\text{f}+\text{d}}$ models. 
	All models are fit on 241 randomly selected cell lines, while performance
	measures are estimated from the remaining 241 cell lines. We repeat every
	simulation Scenario 50 times.

>>>>

	\subsection{Bayesian prior model}\label{sec:prior}
	We carry out inference by endowing the parameters with the following priors:
	\begin{subequations}\label{eq:prior}
		\begin{align}
		  \beta_{jd} | \gamma^2_{jd}, \tau_d^2, \sigma^2_d & \sim \mathcal{N}_{p} 
		  (0, \gamma_{jd}^2 \tau_d^2 \sigma_d^2), 
		  \label{eq:betaprior}\\
		  \gamma_{jd}^{2} & \sim \mathcal{IG}(\phi_{jd}, \lambda_{\text{feat}}), 
		  \label{eq:gammaprior}\\
		  \tau_d^{2} & \sim \mathcal{IG}(\chi_d, \lambda_{\text{drug}}), 
		  \label{eq:tauprior}\\
		  \sigma_d^{2} & \sim 1/\sigma_d^{3}, \label{eq:sigmaprior}
		\end{align}
	\end{subequations}
	where $\mathcal{IG}(\phi, \lambda)$ denotes an inverse Gaussian distribution 
	with mean $\phi$ and shape $\lambda$.

	In model (\ref{eq:prior}), $\gamma_{jd}^2$ in (\ref{eq:gammaprior}) denotes a 
	local variance 
	component that is supposed to capture local, feature-specific variation in the
	model parameters $\beta_{jd}$ in (\ref{eq:betaprior}), 
	while the global variance components $\tau_d^2$ in (\ref{eq:tauprior})
	capture the drug-specific, general trend in $\bm{\beta}_{d}$. Each drug 
	response is endowed with a random error variance $\sigma_d^2$, distributed
	according to (\ref{eq:sigmaprior}).

@@

<<dens_kappa, fig.cap="Implied prior densities $\\pi(\\kappa_{jd})$ for the (a) NIG, (b) Student's $t$, and (c) lasso priors. Different line types correspond to different hyperparameter settings. The hyperparameter settings (given in Section \\ref{sm-sec:hyperparameters} of the SM) were chosen to show some possible, distinct shapes that each of the priors can take.", out.width="100%", fig.asp=1/3>>=
@

  Figure \ref{fig:dens_kappa} shows that, depending on the choice
  of hyperparameters, the NIG prior can behave similarly to the Student's $t$ 
  prior (decreasing form zero, with substantial mass close to zero and 
  little mass close to one, like the
  solid lines in Figure \ref{fig:dens_kappa}a-\ref{fig:dens_kappa}b), 
  but also rather differently (dashed and dotted lines in 
  Figure \ref{fig:dens_kappa}a-\ref{fig:dens_kappa}b). Our argumentation to 
  model the $\gamma^2_{jd}$ by an inverse Gaussian 
  distribution, as has been suggested in \cite{fabrizi_specification_2016} and 
  \cite{caron_sparse_2008}, is three-fold: (i) the NIG model is more flexible 
  than the lasso prior (as seen from Figure 
  \ref{fig:dens_kappa}), (ii) the NIG prior allows to model
  the means of the $\gamma_{jd}^2$ ($\phi_{jd}$) and $\tau_{jd}^2$ 
  ($\chi_{d}$) as a function of external data 
  more conveniently than the Student's $t$ prior, as explained in Section 
  \ref{sec:empiricalbayes}, and (iii) like the horseshoe 
  \cite[]{carvalho_handling_2009}, the NIG shrinkage
  weights prior can put mass both near zero and one, a desirable property of 
  shrinkage priors \cite[]{bernardo_shrink_2011}.

@@

	A pre-processing step selects 100 features with largest variance, while the 
	168 drug sensitivities for 482 cell lines are 
	simulated from model (\ref{eq:likelihood}) and 
	(\ref{eq:prior}). We draw the error
	variances as $\forall d: \sigma_d^2 \sim \Gamma^{-1}(3,2)$,
	such that the prior $\sigma_d^2$ mean and variance are both one. 
	We consider the following three scenarios for the simulation of the 
	drug and feature variance components:
	\begin{itemize}
	  \item{Scenario 1 fixes $\forall d: \tau_d^2=1$ and draws the 
	    $\gamma_{jd}^2$ according to model (\ref{eq:prior}). We
	    create four external dummy feature covariates that code for four 
	    randomly assigned approximately equally sized groups of features. We set
	    $\bm{\alpha}_{\text{feat}}$ such that the $\gamma_{jd}^2$ of the four
	    groups of features have prior means $\phi_{jd} \in \{ 1, 1/2, 1/4, 1/8 \}$
	    ($\bm{\alpha}_{\text{feat}}=\begin{bmatrix} 1 & 1 & 3 & 7 \end{bmatrix} 
	    \tr$). The prior scale parameter is set to $\lambda_{\text{feat}}=1$}.
	  \item{Scenario 2 fixes $\forall j,d: \gamma_{jd}^2=1$ and draws the 
	    $\tau_d^2$ according to model (\ref{eq:prior}), following a procedure
	    similar to the procedure for the $\gamma_{jd}^2$ in Scenario 1: we 
	    randomly create four groups of drugs with corresponding external drug 
	    dummy variables and set $\bm{\alpha}_{\text{drug}}=\begin{bmatrix} 
	    1 & 1 & 3 & 7 \end{bmatrix} \tr$, such that we have $\chi_{d} \in 
	    \{ 1, 1/2, 1/4, 1/8 \}$. The scale is set to $\lambda_{\text{drug}}=1$.}
	  \item{Scenario 3 combines the procedures from Scenarios 1 and 2 to draw both
	    the $\gamma_{jd}^2$ and $\tau_d^2$ according to (\ref{eq:prior}).}
	\end{itemize}
	We estimate four models: (i) the NIG model that only includes an intercept in
	the external covariates, called NIG$_{\text{f}}^-$, NIG$_{\text{d}}^-$, or
	NIG$_{\text{f}+\text{d}}^-$, depending on which variance components are 
	estimated (feature, drug, or both in Scenarios 1, 2, and 3, respectively),
	(ii) the NIG model estimated as in Section \ref{sec:estimation} that includes
	all external covariates, called NIG$_{\text{f}}$, NIG$_{\text{d}}$, or
	NIG$_{\text{f}+\text{d}}$, again depending on which variance components are 
	estimated, and frequentist (iii) lasso and (iv) ridge models.
	Exclusion of the external covariates as in the NIG$_{\text{f}}^-$, 
	NIG$_{\text{d}}^-$, and NIG$_{\text{f}+\text{d}}^-$ models
	amounts to direct estimation of common expected 
	prior means $\phi$ and/or $\chi$, instead of regression estimates for the
	$\phi_{jd}$ and/or $\chi_d$ as in the 
	NIG$_{\text{f}}$, NIG$_{\text{d}}$, and NIG$_{\text{f}+\text{d}}$ models. 
	The lasso and ridge models were fit using the \texttt{R} package 
	\texttt{glmnet} \cite[]{friedman_regularization_2010} with cross validated
	penalty parameters.
	All models are fit on 241 randomly selected cell lines, while performance
	measures are estimated from the remaining 241 cell lines. We repeat every
	simulation Scenario 50 times.

### 2020-03-04 docs/supplement1.Rnw
commit bfeore push

	\section{Jeffrey's prior}
	We parametrise the data with a Gaussian distribution with unknown mean and 
	variance, i.e., $y_i \sim \mathcal{N} (\mu, \sigma^2)$. This gives us the
	following Hessian (matrix of second derivatives) of the log density:
	$$
	\mathbf{H}(\log f(y_i | \mu, \sigma^2)) = \begin{bmatrix}
	-\frac{1}{\sigma^2} & y_i - \mu \\
	y_i - \mu & \frac{1}{2\sigma^4} - \frac{(y_i - \mu)^2}{\sigma^6}.
	\end{bmatrix}
	$$
	Jeffrey's prior is now found as:
	$$
	\pi(\mu, \sigma^2) \propto |\mathcal{I}(\mu, \sigma^2)|^{1/2} = 
	|-\E (\mathbf{H})|^{1/2} = \begin{vmatrix}
	\frac{1}{\sigma^2} & 0 \\
	0 & \frac{1}{2\sigma^4}
	\end{vmatrix}^{1/2} \propto 1/\sigma^3,
	$$
	where $\mathcal{I}(\theta)$ denotes the Fisher information of $\theta$.

>>>>

	\section{Hyperparameter settings}\label{sec:hyperparameters}
	For the NIG shrinkage weights prior in Figure \ref{md-fig:dens_kappa}a the 
	following hyperparameter settings (in terms of the inverse Gaussian prior) 
	were used: $\lambda=1$, $\phi=10$ (solid),
	$\lambda=0.1$, $\phi=10$ (dashed), and $\lambda=1$, $\phi=1$ (dotted).
	The Student`s $t$ prior on the regression parameters $\beta_{jd}$ arises if
	$\gamma_{jd}^2 \sim \Gamma^{-1}(a,b)$, with $\Gamma^{-1}(a,b)$ an inverse
	Gamma distribution with shape $a$ and scale $b$. Figure 
	\ref{md-fig:dens_kappa}b is then obtained by setting: $a=0.9$, $b=1.3$ 
	(solid), $a=1$, $b=4$ (dashed), and $a=1$, $b=0.2$ (dotted).
	The lasso prior results from an exponential prior on $\gamma_{jd}^2$. To
	create Figure \ref{md-fig:dens_kappa}c the following settings were
	used for rate $\lambda$: $\lambda=1$ (solid), $\lambda=0.2$ (dashed), and 
	$\lambda=0.1$ (dotted).

	\section{Jeffrey's prior}\label{sec:jeffreysprior}
	We parametrise the data with a Gaussian distribution with unknown mean and 
	variance, i.e., $y_i \sim \mathcal{N} (\mu, \sigma^2)$. This gives us the
	following Hessian (matrix of second derivatives) of the log density:
	$$
	\mathbf{H}(\log f(y_i | \mu, \sigma^2)) = \begin{bmatrix}
	-\frac{1}{\sigma^2} & y_i - \mu \\
	y_i - \mu & \frac{1}{2\sigma^4} - \frac{(y_i - \mu)^2}{\sigma^6}.
	\end{bmatrix}
	$$
	Jeffrey's prior is now found as:
	$$
	\pi(\mu, \sigma^2) \propto |\mathcal{I}(\mu, \sigma^2)|^{1/2} = 
	|-\E (\mathbf{H})|^{1/2} = \begin{vmatrix}
	\frac{1}{\sigma^2} & 0 \\
	0 & \frac{1}{2\sigma^4}
	\end{vmatrix}^{1/2} \propto 1/\sigma^3,
	$$
	where $\mathcal{I}(\theta)$ denotes the Fisher information of $\theta$.
